---
title: "Introduction and basic workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{roleR_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Motivation

## Installation and setup

```{r setup}
library(roleR)
set.seed(1)
```

## Basic workflow

This is a quick demonstration of the structure of a `roleR` analysis... setting parameters, followed by initializing the model, running the model, optionally creating an experiment, viewing metadata, calculating sumstats... more detail... We will break these down in the following section

... An object of class `roleExperiment` contains multiple different model runs and the parameters used to do those model runs. The parameters are held in an object of class `roleParams` and the model runs are in a list, each element of which is an object of class `roleModel`.

```{r}
library(roleR)

# set-up parameters
p <- roleParams(
  individuals_local = 1000,
  individuals_meta = 10000,
  species_meta = 50,
  speciation_local = 0.5,
  speciation_meta = 1,
  extinction_meta = 0.8,
  trait_sigma = 2,
  env_sigma = 1,
  comp_sigma = 1,
  dispersal_prob = 0.5,
  mutation_rate = 0.01,
  equilib_escape = 1,
  num_basepairs = 250,
  init_type = 'oceanic_island',
  niter = 1000,
  niterTimestep = 100
)

# initialize the model
m <- roleModel(p)

# run the model
m <- runModel(m)

# coerce to an experiment
e <- as(m, 'roleExperiment')

# experiments have a nice data.frame summary of all the params
e@experimentMeta

# we can extract summary statistics; note: rows of summary statistic
# output match rows of `e@experimentMeta`
getSumStats(e, list(rich = richness, hill = hillAbund))
```

## Setting parameters

## Initializing a roleModel

## Running a roleModel

## RoLE Experiments

## Calculating summary statistics

## Analyzing output (machine learning?)
