---
title: "Harmonizing the Lotka-Volterra and MESS models"
author: "A. J. Rominger"
date: "5/25/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Lotka-Volterra competition model

The Lotka-Volterra competition model for two species (here after refered to as LV) looks like this


\begin{align}
\frac{dx_1}{dt} &= r_1 x_1 \left(1 - \frac{x_1 + \alpha_{12} x_2}{K_1} \right) \label{eq:lv1} \\
\frac{dx_2}{dt} &= r_2 x_2 \left(1 - \frac{x_2 + \alpha_{21} x_1}{K_2} \right) \label{eq:lv2}
\end{align}


Our goal is to put this in terms of birth and death rates so we can convert these differential equations to an individual-based model like MESS.

We'll look at just the equation for species 1, recognizing the two equations are the same except for subscripts.  The equation we want is simply the per capita birth rate (times population size) minus the per capita death rate (time population size)


\begin{equation}
\frac{dx_1}{dt} = B(x_1) x_1 - D(x_1) x_1 \label{eq:lvBD}
\end{equation}


Now we need to find meaningful forms for $B(x_1)$ and $D(x_1)$ that are consistent with the LV equations.  A sensible option is 


\begin{align}
B(x_1) &= \lambda_1 \\
D(x_1) &= \mu_1^{(1)} x_1 + \mu_1^{(2)} x_2
\end{align}

These equations say that the per capita birth rate is constant across $x_1$ while the per capita death rate increases linearly with $x_1$ and with $x_2$.

Plugging these form of $B(x_1)$ nd $D(x_1)$ into eq. (\ref{eq:lvBD}) we get


\begin{align}
\frac{dx_1}{dt} &= \lambda_1 x_1 - \mu_1^{(1)} x_1^2 - \mu_1^{(2)} x_1 x_2 
\end{align}


Rearranging eq. (\ref{eq:lv1}) we can quickly see the equivalent terms


\begin{equation}
\frac{dx_1}{dt} = r_1 x_1 - \frac{r_1}{K_1} x_1^2 - \frac{r_1 \alpha_{12}}{K_1} x_1 x_2
\end{equation}


Which gives us


\begin{align}
r_1 &= \lambda_1 \\
\frac{r_1}{K_1} &= \mu_1^{(1)} \\
\frac{r_1 \alpha_{12}}{K_1} &= \mu_1^{(2)}
\end{align}



A few nice things to note: we confirm that $r_1$ is the "rate of increase when rare" (i.e. $\lambda_1$); and we see that "carrying capacity" is inversely proportional to the slopes of the death rates with respect to $x_1$ and $x_2$, i.e., density dependence.

<!-- We can also see that the standard LV parameterization has 3 parameters, while the $B - D$ parameterization has 4 parameters.  Specifically, $\lambda_1^{(0)}$ and $\mu_1^{(0)}$ are not distinguishable, i.e., infinitely many different combinations of $\lambda_1^{(0)}$ and $\mu_1^{(0)}$ will yield the same outcome as long as those combinations obey $r_1 = \lambda_1^{(0)} - \mu_1^{(0)}$.  This will be important to consider when mapping these parameters to the individual-based model. -->


## Taking the LV model to an individual-based framework

To start, let's convert the above birth and death rates to birth and death probabilities.  This is done simply by normalizing the rates


\begin{align}
P_B(x_1) &= \frac{B(x_1)}{B(x_1) + D(x_1)} \\
&= \frac{\lambda_1}{\lambda_1 + \mu_1^{(1)} x_1 + \mu_1^{(2)} x_2} \\
P_D(x_1) &= \frac{D(x_1)}{B(x_1) + D(x_1)} \\
&= \frac{\mu_1^{(1)} x_1 + \mu_1^{(2)} x_2}{\lambda_1 + \mu_1^{(1)} x_1 + \mu_1^{(2)} x_2}
\end{align}


Because we normalized the *per capita* rates, we actually now have the probability of birth or death for one individual of species 1.  That means we've arrived at an individual-based model.


```{r}
#' simulate individual-based lotka-voltera
#' 
#' @param x0 initial condition, should be 2-long int vector
#' @param params a named list of the model params:
#'     * `la1` birth rate for sp 1
#'     * `mu11` intra-specific density dependent death rate for sp 1
#'     * `la2` birth rate for sp 2
#'     * `mu22` intra-specific density dependent death rate for sp 2
#'     * `mu12` density dependent effect of sp 2 on 1
#'     * `mu21` density dependent effect of sp 1 on 2
#' @param nreps single int, the number of timesteps
#' @return matrix of nrow equal to `nreps` and ncol equal to 2 (one for each spp)


lvsim <- function(x0, params, nreps) {
    xt <-  matrix(NA, nrow = nreps, ncol = 2)
    xt[1, ] <- x <- x0
    
    for(b in 2:nreps) {
        # note: current population state is `x`
        
        if(all(x <= 0)) break
        
        # rates
        births <- c(params$la1,
                    params$la2) * x
        deaths <- c(params$mu11 * x[1] + params$mu12 * x[2],
                    params$mu22 * x[2] + params$mu21 * x[1]) * x
        
        # which event happened
        # 1 = sp 1 birth
        # 2 = sp 2 birth
        # 3 = sp 1 death
        # 4 = sp 2 death
        e <- sample(1:4, 1, prob = c(births, deaths))
        
        # update
        if(e <= 2) {
            # birth
            i <- e
            x[i] <- x[i] + 1
        } else {
            # death
            i <- e - 2
            x[i] <- x[i] - 1
        }
        
        xt[b, ] <- x
    }
    
    return(xt)
}


p <- list(la1 = 1,
          mu11 = 0.05,
          la2 = 0.8,
          mu22 = 0.05,
          mu12 = 0.01,
          mu21 = 0.01)

nn <- lvsim(x0 = c(10, 10), params = p, nreps = 1000)

matplot(nn, type = 'l', lty = 1, lwd = 2)
```


To turn this into MESS type IBM we need to implement:

- rocks (aka a null sp that *can* reproduce---a rock reproducing = one of the real spp failing to reproduce); NOTE: but really is that needed? could we have non-reproductive rocks?
- ability to arrive at fully unique and independent la_i and mu_ij....how do we do that with delta, gamma, sigmas, etc? do we need an additional interaction between sigma_c and sigma_e?


## STOP!!!!

let the eq for birth be 

$$
B(x_i) = B_0
$$

and then let death be

$$
D(x_i) = \mu_{0i} + \mu_{ii} x_i + \mu_{ij} x_j
$$

that way $r_i = B_0 - \mu_{0i}$ and we can put all the species-specific stuff in death.  Find $B_0$ by picking $\mu_{01}$ and $\mu_{02}$ such that

$$
r_1 = B_0 - \mu_{01} \\
r_2 = B_0 - \mu_{02} \\
\mu_{01} > 0 \\
\mu_{02} > 0
$$

$$
r_2 = r_1 + \mu_{01} - \mu_{02}
$$


DOES $\mu_{0i}$ matter for death probs?  Seems like $\mu_{0i}$ large would make it so density dependent terms don't matter???

**BELOW IS THE WAY TO GO**

Maybe force $r_1 = B_0$ which makes $\mu_{01} = 0$ and then $\mu{02}$ would have to be whatever satisfies $r_1 = B_0 - \mu_{02}$.  Could pick whether to make $r_1 = B_0$ or $r_2 = B_0$ based on what would keep the mu's positive


Let's re-make the LV function so it takes the classic LV params

```{r}
#' simulate individual-based lotka-voltera
#' 
#' @param x0 initial condition, should be 2-long int vector
#' @param params a named list of the model params:
#'     * `r1` growth rate for sp 1
#'     * `K1` carrying capacity for sp 1
#'     * `alpha12` effect of sp 2 on sp 1
#'     * `r2` growth rate for sp 2
#'     * `K2` carrying capacity of sp 2
#'     * `alpha21` effect of sp 1 on 2
#' @param nreps single int, the number of timesteps
#' @return matrix of nrow equal to `nreps` and ncol equal to 2 (one for each spp)



lvclassic <- function(x0, params, nreps) {
    # re-parameterize
    imax <- which.max(c(params$r1, params$r2))
    B0 <- max(params$r1, params$r2)
    
    mu10 <- ifelse(imax == 1, 0, B0 - params$r1)
    mu20 <- ifelse(imax == 2, 0, B0 - params$r2)
    
    la10 <- params$r1 + mu10
    la20 <- params$r2 + mu20
    
    mu11 <- params$r1 / params$K1
    mu22 <- params$r2 / params$K2
    
    mu12 <- params$r1 * params$alpha12 / params$K1
    mu21 <- params$r2 * params$alpha21 / params$K2
    
    # initialize
    xt <-  matrix(NA, nrow = nreps, ncol = 2)
    xt[1, ] <- x <- x0
    
    for(b in 2:nreps) {
        # note: current population state is `x`
        
        if(all(x <= 0)) break
        
        # rates
        births <- c(la10, la20) * x
        deaths <- c(mu10 + mu11 * x[1] + mu12 * x[2],
                    mu20 + mu22 * x[2] + mu21 * x[1]) * x
        
        # which event happened
        # 1 = sp 1 birth
        # 2 = sp 2 birth
        # 3 = sp 1 death
        # 4 = sp 2 death
        e <- sample(1:4, 1, prob = c(births, deaths))
        
        # update
        if(e <= 2) {
            # birth
            i <- e
            x[i] <- x[i] + 1
        } else {
            # death
            i <- e - 2
            x[i] <- x[i] - 1
        }
        
        xt[b, ] <- x
    }
    
    return(xt)
}


p <- list(r1 = 1.2, 
          K1 = 100, 
          alpha12 = 0.4,
          r2 = 1.1,
          K2 = 120,
          alpha21 = 0.2)

nn <- lvclassic(x0 = c(10, 10), params = p, nreps = 10000)

matplot(nn, type = 'l', lty = 1, lwd = 2)
```


Now make a function where time steps are replacement events (e.g. death then birth)

```{r}
#' simulate individual-based lotka-voltera
#' 
#' @param x0 initial condition, should be 2-long int vector
#' @param params a named list of the model params:
#'     * `r1` growth rate for sp 1
#'     * `K1` carrying capacity for sp 1
#'     * `alpha12` effect of sp 2 on sp 1
#'     * `r2` growth rate for sp 2
#'     * `K2` carrying capacity of sp 2
#'     * `alpha21` effect of sp 1 on 2
#' @param nreps single int, the number of timesteps
#' @return matrix of nrow equal to `nreps` and ncol equal to 2 (one for each spp)



lvrep <- function(x0, params, nreps) {
    # re-parameterize
    imax <- which.max(c(params$r1, params$r2))
    B0 <- max(params$r1, params$r2)
    
    mu10 <- ifelse(imax == 1, 0, B0 - params$r1)
    mu20 <- ifelse(imax == 2, 0, B0 - params$r2)
    
    la10 <- params$r1 + mu10
    la20 <- params$r2 + mu20
    
    mu11 <- params$r1 / params$K1
    mu22 <- params$r2 / params$K2
    
    mu12 <- params$r1 * params$alpha12 / params$K1
    mu21 <- params$r2 * params$alpha21 / params$K2
    
    J <- params$K1 + params$K2
    x0 <- c(x0, J - sum(x0)) # add "rocks"
    
    # initialize
    xt <-  matrix(NA, nrow = nreps, ncol = 3)
    xt[1, ] <- x <- x0
    
    for(b in 2:nreps) {
        # note: current population state is `x`
        
        if(all(x <= 0)) break
        
        # death rates
        deaths <- c(mu10 + mu11 * x[1] + mu12 * x[2],
                    mu20 + mu22 * x[2] + mu21 * x[1], 
                    1) * x
        
        # who dies
        idead <- sample(1:3, 1, prob = deaths)
        x[idead] <- x[idead] - 1
        
        # birth rates
        births <- c(la10, la20, 1) * (x + c(0, 0, 1)) # the added one 
                                                      # incorporates the extra 
                                                      # immigration prob
        
        # who replaces
        iborn <- sample(1:3, 1, prob = births)
        x[iborn] <- x[iborn] + 1
        
        # update
        xt[b, ] <- x
    }
    
    return(xt)
}


p <- list(r1 = 1.2, 
          K1 = 100, 
          alpha12 = 0.4,
          r2 = 1.1,
          K2 = 120,
          alpha21 = 0.2)

nnZS <- lvrep(x0 = c(10, 10), params = p, nreps = 10000)
nnLV <- lvclassic(x0 = c(10, 10), params = p, nreps = 10000)

par(mfcol = c(2, 1), mar = c(2, 2, 0, 0) + 0.5)
matplot(nnLV[, 1:2], type = 'l', lty = 1, lwd = 2)
matplot(nnZS[, 1:2], type = 'l', lty = 1, lwd = 2)

```



The raw timing is clearly different, but let's see if equilib is the same.  We'll say equilib is achieved post 4000 iterations for both versions

```{r}
p <- list(r1 = 1.4, 
          K1 = 180, 
          alpha12 = 0.4,
          r2 = 1.1,
          K2 = 100,
          alpha21 = 0.1)

sims <- parallel::mclapply(1:500, mc.cores = 10, FUN = function(i) {
    nnZS <- lvrep(x0 = c(10, 10), params = p, nreps = 10000)[-(1:4000), 1:2]
    nnLV <- lvclassic(x0 = c(10, 10), params = p, nreps = 10000)[-(1:4000), ]
    
    return(c(colMeans(nnZS), colMeans(nnLV)))
})

sims <- do.call(rbind, sims)

d <- lapply(1:ncol(sims), function(i) {
    thisD <- density(sims[, i])
    o <- as.data.frame(thisD[c('x', 'y')])
    
    return(cbind(ID = i, o))
})

d <- do.call(rbind, d)

plot(d$x, d$y, col = c('red', 'blue', 'orange', 'darkslateblue')[d$ID])
```



We want a death and birth every time step...how to do that?  Scale it so $P_D$ and $P_B$ equal 1 at equilibirum?  Scale the $K_i$'s to equal $J$?

If we do that, I think we get conversion to MESS params for free 



## The MESS model

The MESS model is individual-based.  Each time step an individual dies and is immediately replaced with a new individual. We will start by considering the MESS model for only 2 species, and with no immigration, speciation, or trait evolution.  Thus, death events are replaced only by births.  Species 1 and 2 each have only one distinct trait value (call them $t_1$ and $t_2$).  There is also an environmental optimum, call it $\hat{t}$.  There are three mechanisms for death: neutral, trait-based competition, environmental filtering.  For **one individual** belonging to species 1, the rates of death from **n**eutral, **c**ompetitive, and **e**nvironmental filtering processes are



\begin{align}
D_N(n_1) &= 1 \\
D_C(n_1) &= \frac{1}{J} \sum_{j = 1}^J \exp \left[ \frac{1}{\sigma_C} (t_1 - t_j)^2 \right] \\
D_E(n_1) &= 1 - \exp \left[ \frac{1}{\sigma_E} (t_1 - \hat{t})^2 \right]
\end{align}

The *probability* of death in one time step for this individual is

\begin{equation}
P_D(n_1) = \delta D_N(n_1) + (1 - \delta) (\gamma D_C + (1 - \gamma) D_E)
\end{equation}


The deth rate due to competition can be further 





