---
title: "roleR demonstration"
format: html
---

## Set-up

```{r}
library(roleR)
```

## Parameterization

We make a parameters object

```{r}
p <- roleParams()
```


## Making and running a model

First we make the model object

```{r}
mod <- roleModel(p)
```

Design decisions:

- mostly only expose "observable" data like IDs of individuals (do not show things like times of local extinctions)
- one exception is that we do show time of last origin because time of local duration is interesting
- do not show aggregate statistics like do not aggregate spp abundances from IDs of individuals 

Now we can run it

```{r}
modRun <- runRole(mod)
```


Issues:

- code in `sim_oo.cpp` assumes params will be S4 `roleParams` not a list, and also that params will be constant, not time variable
- could perhaps make the time-variable params into a matrix instead of list, and pass param values one row at a time???
- or have `getParamDouble` also take an index of which iteration weʻre on and thus which element to extract
